Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-04-05 paper/paper.tex
More data
More things

\usepackage{microtype}                 % typographic perfection
\usepackage{complexity}                % \P, \NP etc
\usepackage{tikz}                      % For pretty pictures
\usepackage{amsmath}                   % \operatorname
\usepackage{amsfonts}                  % \mathbb
\usepackage{hyperref}                  % clicky links
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{gnuplot-lua-tikz}          % graphs
\usepackage[ruled,vlined]{algorithm2e} % algorithms (after cleverref!)
\usepackage{cite}                      % automatically sort mass cites
\usepackage[bottom]{footmisc}          % fix footnote + bottom figure on same page

@@

@@

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two approaches using modern algorithms. We also investigate a
    restricted branching approach for the connected version of the problem, show how to implement
    this in constraint-based and clique-inspired algorithms, and compare it to conventional
    propagation.
\end{abstract}

@@

We will be working with graphs which may be directed, which may have labels associated with
vertices, edges, or both, and which may contain loops. To simplify notation, we will work with a
single definition throughout: we define a \emph{graph} $G = (V, \ell, e)$ to be a set of vertices
$V$, a vertex labelling function $\ell : V \rightarrow \mathbb{N}$, and an edge labelling function
$e : V \times V \rightarrow \mathbb{N} \cup \{ \bot \}$. To represent graphs without vertex (or
edge) labels, we just pick our favourite number and give it to every vertex (or edge). We also use a
distinct label $\bot$ to represent the \emph{lack of} an edge between two vertices.  We write
$\operatorname{V}(G)$ for $V$, and $\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood}
of a vertex $v$ (that is, the set of vertices $w$ such that either $e(v, w) \ne \bot$ or $e(w, v)
\ne \bot$); a vertex is \emph{isolated} if its neighbourhood is empty. A \emph{path} between two
vertices $v$ and $w$ is a sequence of $n$ vertices $v = v_1, v_2, \ldots, v_n = w$ such that for
each $m \in \{ 1 \ldots n - 1 \}$, $e(v_m, v_{m + 1}) \ne \bot$.  A graph is \emph{connected} if a
path exists between each distinct pair of vertices.

@@

Another widely studied approach is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover) in an association graph (which we describe
below). Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

(Other approaches have been tried, including mixed integer programming
\cite{DBLP:journals/anor/PivaS12}; SAT encodings seem to struggle even for subgraph isomorphism??
cite IJCAI if it's not rejected?.)

@@

@@

@@

\begin{figure}[p]
    \newcounter{stepcounter2}
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \matrix (m) [matrix of nodes, column sep=0.5cm] {
            \begin{tikzpicture}[scale=0.33]
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Ma) at (0, 4) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mb) at (2, 4) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mc) at (0, 2) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Md) at (2, 2) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Me) at (0, 0) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mf) at (2, 0) {\phantom{0}};

@@

Finally, we describe the colouring process. In conventional clique algorithms, a simple greedy
sequential colouring is used (possibly with the help of previous colourings to reduce the
computational cost \cite{DBLP:conf/lion/NikolaevBS15}, and possibly with shortcuts taken for certain
vertices \cite{DBLP:journals/cor/SegundoT14}, and possibly followed by a repair step to
improve the colouring \cite{DBLP:conf/walcom/TomitaSHTW10}, or stronger bounding rules based upon
MaxSAT inference \cite{DBLP:conf/ictai/LiFX13,DBLP:conf/lion/LiJX15,DBLP:journals/cor/SegundoNB15}).
Such colourings will not give us the required property that vertices in $\mathit{remaining} \cap
\mathit{connected}$ come last (so they are selected first by the reverse branching order). Thus we
produce two greedy sequential colourings, first considering the non-branching vertices in
$\mathit{remaining} \setminus \mathit{connected}$, followed by the branching vertices, and
concatenate them (\lineref{makecolours}). This produces a valid colouring, since we do not merge any
colour classes between the two stages, although it may use more colours than a single colouring
would.

Our $\FuncSty{colour}$ procedure is the same as the bit-parallel algorithm introduced by San Segundo
\textit{et al.}\ \cite{DBLP:journals/cor/SegundoRJ11}. While we have remaining vertices to colour
(\lineref{whileuncoloured}), we start a new colour class (\lineref{newcolourclass}), and then
repeatedly pick a legal vertex to add to that colour class (\lineref{addtocolourclass}).  The
selection of the next vertex to colour (\lineref{selectvtocolour}) may be performed efficiently in
hardware if the association graph is permuted to be in degree order at the top of search. (Other
initial vertex orderings have been considered on general clique problems
\cite{DBLP:journals/algorithms/Prosser12,DBLP:conf/lion/SegundoLB14}; it is possible that special
properties of the association graph could be exploited in this step.)

@@

@@

@@

\bibliographystyle{splncs}
\bibliography{paper}

>>>>

\usepackage{microtype}                 % typographic perfection
\usepackage{complexity}                % \P, \NP etc
\usepackage{tikz}                      % For pretty pictures
\usepackage{amsmath}                   % \operatorname
\usepackage{amsfonts}                  % \mathbb
\usepackage{hyperref}                  % clicky links
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{gnuplot-lua-tikz}          % graphs
\usepackage[ruled,vlined]{algorithm2e} % algorithms (after cleverref!)
\usepackage{cite}                      % automatically sort mass cites
\usepackage[bottom]{footmisc}          % fix footnote + bottom figure on same page
\usepackage{placeins}

@@

\renewcommand{\floatpagefraction}{0.8}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}

@@

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two approaches using modern algorithms, and see that the best
    choice depends mainly upon whether the graphs have labelled edges. We also investigate a
    restricted branching approach for the connected version of the problem, show how to implement
    this in constraint-based and clique-inspired algorithms, and compare it to conventional
    propagation.
\end{abstract}

@@

We will be working with graphs which may be directed, which may have labels associated with
vertices, edges, or both, and which may contain loops. To simplify notation, we will work with a
single definition throughout: we define a \emph{graph} $G = (V, \ell, e)$ to be a set of vertices
$V$, a vertex labelling function $\ell : V \rightarrow \mathbb{N}$, and an edge labelling function
$e : V \times V \rightarrow \mathbb{N} \cup \{ \bot \}$. To represent graphs without vertex (or
edge) labels, we just pick our favourite number and give it to every vertex (or edge). We also use a
distinct label $\bot$ to represent the \emph{lack of} an edge between two vertices; vertices $v$ and
$w$ are said to be \emph{adjacent} if $e(v, w) \ne \bot$ or $e(w, v) \ne \bot$. We write
$\operatorname{V}(G)$ for $V$, and $\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood}
of a vertex $v$ (that is, the set of vertices adjacent to $v$); a vertex is \emph{isolated} if its
neighbourhood is empty. A \emph{path} between two vertices $v$ and $w$ is a sequence of $n$ vertices
$v = v_1, v_2, \ldots, v_n = w$ such that for each $m \in \{ 1 \ldots n - 1 \}$, $v_m$ is adjacent
to $v_{m + 1}$. A graph is \emph{connected} if a path exists between each distinct pair of vertices.

@@

Another widely studied approach is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph (which we describe
below). Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

(Other approaches have been tried, including dedicated algorithms ?? cite, mixed integer programming
\cite{DBLP:journals/anor/PivaS12}, and heuristics ?? cite; SAT encodings seem to struggle even for
subgraph isomorphism??  cite IJCAI.)

@@

% \begin{figure}[p]
%     \centering
%     \input{gen-graph-unconnected-33-breakdown}
%     \caption{Clique versus constraint programming, 33\% labelled, broken down by family.}
%         \label{figure:unconnected-33-breakdown}
% \end{figure}

@@

% \begin{figure}[p]
%     \centering
%     \input{gen-graph-unconnected-plain-breakdown}
%     \caption{Clique versus constraint programming, undirected, unlabelled, broken down by family.}
%         \label{figure:unconnected-plain-breakdown}
% \end{figure}

@@

\begin{figure}[tb]
    \newcounter{stepcounter2}
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \matrix (m) [matrix of nodes, column sep=0.5cm] {
            \begin{tikzpicture}[scale=0.33]
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Ma) at (0, 4) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mb) at (2, 4) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mc) at (0, 2) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Md) at (2, 2) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Me) at (0, 0) {\phantom{0}};
                    \node[draw, circle, fill=white, inner sep=0.5pt, font=\normalsize] (Mf) at (2, 0) {\phantom{0}};

@@

Finally, we describe the colouring process. In conventional clique algorithms, a simple greedy
sequential colouring is used (possibly with the help of previous colourings to reduce the
computational cost \cite{DBLP:conf/lion/NikolaevBS15}, and possibly with shortcuts taken for certain
vertices \cite{DBLP:journals/cor/SegundoT14}, and possibly followed by a repair step to
improve the colouring \cite{DBLP:conf/walcom/TomitaSHTW10}, or stronger bounding rules based upon
MaxSAT inference \cite{DBLP:conf/ictai/LiFX13,DBLP:conf/lion/LiJX15,DBLP:journals/cor/SegundoNB15}).
Such colourings will not give us the required property that vertices in $\mathit{remaining} \cap
\mathit{connected}$ come last (so they are selected first by the reverse branching order). Thus we
produce two greedy sequential colourings, first considering the non-branching vertices in
$\mathit{remaining} \setminus \mathit{connected}$, followed by the branching vertices, and
concatenate them (\lineref{makecolours}). This produces a valid colouring, since we do not merge any
colour classes between the two stages, although it may use more colours than a single colouring
would\footnote{What if we did not guarantee that vertices in $\mathit{remaining} \cap
    \mathit{connected}$ came last, and just used a conventional colouring with the branching rule?
    Suppose we had four vertices in $\mathit{remaining}$, and produced a colouring $[[v_1, v_2],
    [v_3], [v_4]]$, and suppose that extending $\mathit{solution}$ with $\{ v_1, v_3, v_4 \}$ gives
an optimal solution. If $v_4$ was not $\mathit{connected}$ yet, we would not branch on that subtree,
and the bound could eliminate branching on $v_3$ and $v_1$, so we would miss the solution. Thus we
cannot simply add the branching rule without also adapting the combined bound / ordering heuristic.}.

Our $\FuncSty{colour}$ procedure is the same as the bit-parallel algorithm introduced by San Segundo
\textit{et al.}\ \cite{DBLP:journals/cor/SegundoRJ11}. While we have remaining vertices to colour
(\lineref{whileuncoloured}), we start a new colour class (\lineref{newcolourclass}), and then
repeatedly pick a legal vertex to add to that colour class (\lineref{addtocolourclass}).  The
selection of the next vertex to colour (\lineref{selectvtocolour}) may be performed efficiently in
hardware if the association graph is permuted to be in degree order at the top of search. Other
initial vertex orderings have been considered on general clique problems
\cite{DBLP:journals/algorithms/Prosser12,DBLP:conf/lion/SegundoLB14}; it is possible that special
properties of the association graph could be exploited in this step.

@@

% \begin{figure}[p]
%     \centering
%     \input{gen-graph-connected-undir33-breakdown}
%     \caption{Clique versus constraint programming, connected, undirected, 33\% labelled, broken down by family.}
%         \label{figure:connected-undir33-breakdown}
% \end{figure}

\begin{figure}[tb]
    \centering
    \input{gen-graph-connected-cp-undir33.tex}
    \caption{A pairwise comparison of different CP techniques for the connected, undirected, 33\%
        labelled variant with graphs of up to 80 vertices. The colour at a point indicates the
        number of instances with these runtimes, with darker colours representing more instances.}
        \label{figure:connected-cp-undir33}
\end{figure}

@@

% \begin{figure}[p]
%     \centering
%     \input{gen-graph-connected-plain-breakdown}
%     \caption{Clique versus constraint programming, connected, undirected, unlabelled, broken down by family.}
%         \label{figure:connected-plain-breakdown}
% \end{figure}

\begin{figure}[tb]
    \centering
    \input{gen-graph-connected-cp-plain.tex}
    \caption{Different CP techniques, connected, undirected, unlabelled, up to 35 vertices.}
        \label{figure:connected-cp-plain}
\end{figure}

@@

\FloatBarrier
\bibliographystyle{splncs}
\bibliography{paper}
